# Community composition

## Filter data

```{r load_data_community, comment="", message=FALSE, warning=FALSE, eval=FALSE}
load("data/data.Rdata")
```

Filter samples with high host data
```{r load_data_host_filtering, comment="", message=FALSE, warning=FALSE, eval=FALSE}
genome_counts_filt <- genome_counts %>%
  select(one_of(c("genome",sample_metadata$sample)))%>%
  filter(rowSums(. != 0, na.rm = TRUE) > 0) %>%
  select_if(~!all(. == 0))
genome_counts <- genome_counts_filt
genome_metadata <- genome_metadata %>% 
  semi_join(., genome_counts_filt, by = "genome") %>% 
  arrange(match(genome,genome_counts_filt$genome))

genome_tree <- keep.tip(genome_tree, tip=genome_metadata$genome) # keep only MAG tips

#load("data/genome_gifts.Rdata")
```

Make a phyloseq object

```{r phyloseq_objects1, warning=FALSE, comments="", message=FALSE, eval=FALSE}
phylo_samples <- sample_metadata %>% 
  column_to_rownames("sample") %>% 
  sample_data() #convert to phyloseq sample_data object
phylo_genome <- genome_counts_filt %>% 
  column_to_rownames("genome") %>% 
  otu_table(., taxa_are_rows = TRUE)
phylo_taxonomy <- genome_metadata %>%
  column_to_rownames("genome") %>% 
  as.matrix() %>% 
  tax_table() #convert to phyloseq tax_table object
phylo_tree <- phy_tree(genome_tree) 

physeq_genome <- phyloseq(phylo_genome, phylo_taxonomy, phylo_samples,phylo_tree)
physeq_genome_clr <- microbiome::transform(physeq_genome, 'clr')
```

```{r wrap_working_objects_filtered, warning=FALSE, comments="", message=FALSE, eval=FALSE}
save(sample_metadata, 
     genome_metadata, 
     read_counts, 
     genome_counts, 
     genome_counts_filt, 
     genome_tree,
     physeq_genome,
     physeq_genome_clr,
     genome_gifts_raw, 
#     genome_gifts,
     phylum_colors,
     file = "data/data_host_filtered.Rdata")
```


## Load data

```{r load_data_host_filtered, comment="", message=FALSE, warning=FALSE}
load("data/data_host_filtered.Rdata")
```

## Taxonomy overview 

### Stacked barplot

```{r taxonomy_barplot, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS normalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  left_join(., sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  filter(count > 0) %>% #filter 0 counts
  ggplot(., aes(x=sample,y=count, fill=phylum, group=phylum)) + #grouping enables keeping the same sorting of taxonomic units
    geom_bar(stat="identity", colour="white", linewidth=0.1) + #plot stacked bars with white borders
    scale_fill_manual(values=phylum_colors) +
 #   facet_nested(. ~ region+diet,  scales="free") + #facet per day and treatment
    guides(fill = guide_legend(ncol = 1)) +
    theme(
          axis.title.x = element_blank(),
          panel.background = element_blank(),
          panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          strip.background = element_rect(fill = "white"),
    strip.text = element_text(size = 12, lineheight = 0.6),
    strip.placement = "outside",
    axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black")) +
   labs(fill="Phylum",y = "Relative abundance",x="Samples")
```

**Number of bacteria phyla**

```{r phyla, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

**Bacteria phyla in wild individuals**

```{r phyla_nat, comment="", echo=FALSE, message=FALSE, warning=FALSE}

wild_samples <- sample_metadata %>% 
  filter(region=="Gipuzkoa") %>% 
  dplyr::select(sample) %>% 
  pull()

wild_genomes <- genome_counts %>% 
  column_to_rownames("genome") %>% 
  select(all_of(wild_samples)) %>%
  as.data.frame() %>%
  filter(rowSums(across(where(is.numeric)))!=0)%>% 
  rownames_to_column("genome")%>% 
  dplyr::select(genome) %>% 
  pull()

genome_metadata %>% 
  filter(genome %in% wild_genomes) %>% 
  filter(domain == "d__Bacteria")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length() 
```


**Number of Archaea phyla**

```{r arch, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>% 
  filter(domain == "d__Archaea")%>%
  dplyr::select(phylum) %>%
  unique() %>%
  pull() %>%
  length()
```

### Genus and species annotation

**Number of MAGs without species-level annotation**
```{r nonspe, comment="", echo=FALSE, message=FALSE, warning=FALSE}
genome_metadata %>%
  filter(species == "s__") %>%
  summarize(Mag_nospecies = n())%>%
  select(Mag_nospecies) %>% 
  pull()

```

```{r nonspe_phylum, comment="", echo=FALSE, message=FALSE, warning=FALSE}
total_mag_phylum <- genome_metadata %>%
  group_by(phylum) %>%
  summarize(count_total = n())
genome_metadata %>%
  filter(species == "s__") %>%
  group_by(phylum) %>%
  summarize(count_nospecies = n()) %>% 
  left_join(total_mag_phylum, by = join_by(phylum == phylum)) %>% 
  mutate(percentage=100*count_nospecies/count_total) %>% 
  tt()

```

**Percentage of MAGs without species-level annotation**
```{r sp_percet, comment="", echo=FALSE, message=FALSE, warning=FALSE}
nmags <- nrow(genome_counts)
nonspecies <- genome_metadata %>%
    filter(species == "s__") %>%
    nrow()
perct <- nonspecies*100/nmags
perct
```

**Number of MAGs without genera-level annotation**
```{r nongenera, comment="", echo=FALSE, message=FALSE, warning=FALSE}
nongenera <- genome_metadata %>%
    filter(genus == "g__") %>%
    nrow()
cat(nongenera)
```


### Phylum relative abundances

```{r taxonomy_phylum_summary, warning=FALSE, comments="", message=FALSE}
phylum_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>%
  left_join(sample_metadata, by = join_by(sample == sample)) %>%
  left_join(genome_metadata, by = join_by(genome == genome)) %>%
  group_by(sample,phylum,region) %>%
  summarise(relabun=sum(count))
```

```{r taxonomy_boxplot_phylum, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
phylum_arrange <- phylum_summary %>%
    group_by(phylum) %>%
    summarise(mean=mean(relabun)) %>%
    arrange(-mean) %>%
    select(phylum) %>%
    pull()

phylum_summary %>%
    filter(phylum %in% phylum_arrange) %>%
    mutate(phylum=factor(phylum,levels=rev(phylum_arrange))) %>%
    ggplot(aes(x=relabun, y=phylum, group=phylum, color=phylum)) +
        scale_color_manual(values=phylum_colors[rev(phylum_arrange)]) +
        geom_jitter(alpha=0.5) + 
        theme_minimal() + 
        theme(legend.position="none") +
        labs(y="Phylum",x="Relative abundance")
```

```{r taxonomy_jitterplot_phylum, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
phylum_arrange <- phylum_summary %>%
    group_by(phylum) %>%
    summarise(mean=sum(relabun)) %>%
    arrange(-mean) %>%
    select(phylum) %>%
    pull()

phylum_summary %>%
    left_join(genome_metadata %>% select(phylum,phylum) %>% unique(),by=join_by(phylum==phylum)) %>%
#    left_join(sample_metadata,by=join_by(sample==sample)) %>%
    filter(phylum %in% phylum_arrange[1:20]) %>%
    mutate(phylum=factor(phylum,levels=rev(phylum_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=phylum, group=phylum, color=phylum)) +
        scale_color_manual(values=phylum_colors[-8]) +
        geom_jitter(alpha=0.5) + 
        theme_minimal() + 
        labs(y="phylum", x="Relative abundance", color="Phylum")
```

## Taxonomy boxplot

### Family

```{r taxonomy_family_summary, warning=FALSE, comments="", message=FALSE}
family_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,family,region) %>%
  summarise(relabun=sum(count))
```

```{r taxonomy_jitterplot_family, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
family_arrange <- family_summary %>%
    group_by(family) %>%
    summarise(mean=sum(relabun)) %>%
    arrange(-mean) %>%
    select(family) %>%
    pull()

family_summary %>%
    left_join(genome_metadata %>% select(family,phylum) %>% unique(),by=join_by(family==family)) %>%
    filter(family %in% family_arrange[1:20]) %>%
    mutate(family=factor(family,levels=rev(family_arrange[1:20]))) %>%
    filter(relabun > 0) %>%
    ggplot(aes(x=relabun, y=family, group=family, color=phylum)) +
        scale_color_manual(values=phylum_colors[-8]) +
        geom_jitter(alpha=0.5) + 
        theme_minimal() + 
        labs(y="Family", x="Relative abundance", color="Phylum")
```

### Genus

```{r taxonomy_genus_summary, warning=FALSE, comments="", message=FALSE}
genus_summary <- genome_counts_filt %>%
  mutate_at(vars(-genome),~./sum(.)) %>% #apply TSS nornalisation
  pivot_longer(-genome, names_to = "sample", values_to = "count") %>% #reduce to minimum number of columns
  left_join(sample_metadata, by = join_by(sample == sample)) %>% #append sample metadata
  left_join(genome_metadata, by = join_by(genome == genome)) %>% #append genome metadata
  group_by(sample,phylum,genus) %>%
  summarise(relabun=sum(count)) %>%
  filter(genus != "g__") %>%
  mutate(genus= sub("^g__", "", genus))
```


